**Title:** Analyzing long-term water quality of lakes in Rhode Island and the Northeastern United States with an anomaly approach

### 2021-03-19 manuscript/ecosphere_production/manuscript.Rmd
figures and tables and citations, oh my!

---
title: "**Title:** Analyzing long-term water quality of lakes in Rhode Island and the Northeastern United States with an anomaly approach"
author:
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '**Authors:** Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. D. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A. ^*2*^'
- '**Corresponding author: hollister.jeff@epa.gov**'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: |
  | **Abstract:** Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management. Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets. Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured. We use the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset to determine if there are sub-regional (i.e. 3000km^2^) water quality trends. URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional scales, respectively. We assess regional (LAGOSNE) and sub-regional (URIWW) trends with yearly median anomalies calculated on a per-station basis. Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016. Total nitrogen, total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable. At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, and N:P all stable over time.  Total phosphorus does show a very slight increase. In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, despite stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio. Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends. 
  | **Keywords:** Volunteer Monitoring, community Science, Nutrients, Lakes, Trend Analysis, Long Term
---

@@

The goals of this study were to examine ~25 years of lake and reservoir data in Rhode Island and answer two questions.  First, are there state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature? Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United States?  Another objective of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends. We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.4306765> [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation; @sorannolimno2019].

@@

## Study Area and Data

@@

Prior studies have modeled water quality trends across a larger region of the northeastern US within 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017]. We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al. [-@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017; @sorannolimno2019]. Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.  We used the same selection criteria on the LAGOSNE dataset as was applied to the URIWW data. This ensured that both datasets represented the same seasonal time frame.  Exact measurement depth for the LAGOSNE data were not available; however, the water quality data are reported to be either surface or epilimnion samples [@sorannolagosne2017]. At the time of our analysis, the LAGOSNE data do not extend beyond 2013 thus sites needed to have a sampling event in 1993-2002 and in 2003-2013 to be included.

@@

Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4306765> [@zenodo_citation].  All analyses were conducted with R version 4.0.3 and details on R package versions and operating system used for this analysis are included in a file, `sessioninfo.txt` at <https://github.com/USEPA/ri_wq_trends/blob/master/sessioninfo.txt> [@r2020; @lubridate; @broom; @stringr; @dplyr; @readr; @readxl; @tidyr; @here; @ggplot2; @hrbrthemes; @kendall; @sf; @usaboundaries; @usaboundariesdata; @knitr; @captioner; @pander; @cowplot; @tidycensus; @feddata; @nsink; @units; @xie2015dynamicdoc].  Values included in each of the figures have also been separately saved as a comma-separated values file, `yearly_average_anomaly.csv`, and may be retrieved via <https://github.com/USEPA/ri_wq_trends/blob/master/data/yearly_average_anomaly.csv>.

## Water Quality Trend Analysis

@@

Median annual temperature anomalies in lakes and reservoirs appear to be increasing as the slope is greater than the threshold of 0.02 we identified, the p-value suggests some statistical support for a positive trend (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`), and the majority of years with median temperature greater than the long-term median are occurring in the second half of the time period (`r fig_num(name = "temp_ri", caption = "Twenty year trend for median lake temperature anomaly in Rhode Island. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians. Black line is the linear fit for year and median anomaly.", display = "cite")`, `r tab_num(name = "trend_summ", caption = 'Summary of long term water quality trends in the URIWW and LAGOSNE data. The "++" indicate a positive trend, the "+" indicates a weak positive trend, and the "0" indicates no trend.', display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`).  The slope of `r signif(ww_chla_rl$slope,2)` μg/L is greater than our threshold of 0.035 μg/L and the above-average years have mostly occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Twenty year trend for median chlorophyll anomaly.  Panel A. URI Watershed Watch yearly chlorophyll anomalies.  Panel B. LAGOSNE yearly chlorophyll anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians. Black line is the linear fit for year and median anomaly.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).

@@

In general, there was little evidence to suggest broad regional changes in chlorophyll *a* as it showed a very weak positive trend, slight statistical support and above average years spread evenly throughout the time period (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Total nitrogen showed a slight decreasing trend with weak statistical support and slope less than the 3.5 μg/L that would result in a trophic state change over a century, thus there is little support for a meaningful trend in TN at the regional scale (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Furthermore, The last two years for which LAGOSNE had data on total nitrogen were from a relatively smaller number of sites (`r fig_num("year_n", display = "cite")`) and those values may not be representative.  Total phosphorus showed a very small increasing trend and there is statistical support for that trend.  The slope however suggests little support for an ecological meaningful change in total phosphorus (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`). Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`) as none of our criteria for a trend were met.  Although, there appears to be a non-linear trend that suggests increasing TN:TP early in the time period and decreasing TN:TP later, we feel we cannot effectively evaluate this though as total nitrogen measurements were rare in 2011 (`r fig_num("year_n", display = "cite")`) and sites that had both nitrogen and phosphorus were even rarer.  This low median anomaly should be interpreted with this caveat in mind.   Taken together we feel our results largely match the findings of Oliver et al. [-@Oliver_2017] that there is stasis in lake nutrients and chlorophyll within the LAGOSNE region.

@@

```{r context, echo=FALSE, results="asis"}
#knitr::kable(context_summ, format = "pipe")
knitr::kable(context_summ)
```
`r tab_num("context_summ")`

@@

```{r ww_params, echo=FALSE, results="asis"}
knitr::kable(ww_param_ranges, format = "pipe")
```
`r tab_num("ww_params")`

@@

```{r lagos_params, echo=FALSE, results="asis"}
knitr::kable(lagos_param_ranges, format = "pipe")
```
`r tab_num("lagos_params")`

@@

```{r ts_table, echo=FALSE, results="asis"}
knitr::kable(ts_table, format = "pipe")
```
`r tab_num("ts_table")`

@@

trend_summ_df <- tibble(Source = c("URIWW", "LAGOSNE"), 
                        Temperautre = c("++", ""), Chlorophyll= c("++", "0"), 
                        `Total Nitrogen` = c("0", "0"), 
                        `Total Phosphorus`= c("0", "+"), 
                        `Nitrogen:Phosphorus` = c("0", "0"))

knitr::kable(trend_summ_df, format = "pipe")
```
`r tab_num("trend_summ")`

>>>>

---
title: "**Title:** Analyzing long-term water quality of lakes in Rhode Island and the Northeastern United States with an anomaly approach"
author:
- '**List of Authors:** Hollister. J. W. ^†^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. D. ^*1*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A. ^*2*^'  
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- '**†: hollister.jeff@epa.gov**'
- "**Track:** Freshwater Ecology"
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: |
  | **Abstract:** Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management. Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets. Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured. We used the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset to determine if there were sub-regional (i.e. 3000km^2^) water quality trends. URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. The LAGOSNE and URIWW datasets allowed for comparison of water quality trends at regional and sub-regional scales, respectively. We assessed regional (LAGOSNE) and sub-regional (URIWW) trends with yearly median anomalies calculated on a per-station basis. Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016. Total nitrogen, total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable. At the regional scale, the LAGOSNE dataset showed similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, and N:P all stable over time.  Total phosphorus did show a very slight increase. In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs increased, despite stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio. Additionally, we demonstrated both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends. 
  | **Keywords:** Anomaly; Chlorophyll; Community Science; Lakes; Lake Temperature; Long Term; Nutrients; Trend Analysis; Volunteer Monitoring.  
---

@@

The goals of this study were to examine ~25 years of lake and reservoir data in Rhode Island and answer two questions.  First, are there state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature? Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United States?  Another objective of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends. We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.4306765> [@sorannolagosne2017; @stacheleklagosne; @zenodo_citation; @sorannolimno2019].

@@

## Study area and data

@@

Prior studies have modeled water quality trends across a larger region of the northeastern US within 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017]. We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al. [-@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne; @sorannolagosne2017; @sorannolimno2019]. Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.  We used the same selection criteria on the LAGOSNE dataset as was applied to the URIWW data. This ensured that both datasets represented the same seasonal time frame.  Exact measurement depth for the LAGOSNE data were not available; however, the water quality data are reported to be either surface or epilimnion samples [@sorannolagosne2017]. At the time of our analysis, the LAGOSNE data do not extend beyond 2013 thus sites needed to have a sampling event in 1993-2002 and in 2003-2013 to be included.

@@

Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4306765> [@zenodo_citation].  All analyses were conducted with R version 4.0.3 and details on R package versions and operating system used for this analysis are included in a file, `sessioninfo.txt` at <https://github.com/USEPA/ri_wq_trends/blob/master/sessioninfo.txt> [@r2021; @lubridate; @broom; @stringr; @dplyr; @readr; @readxl; @tidyr; @here; @ggplot2; @hrbrthemes; @kendall; @sf; @usaboundaries; @usaboundariesdata; @knitr; @captioner; @pander; @cowplot; @tidycensus; @feddata; @nsink; @units; @xie2015dynamicdoc; @magick; @flextable].  Values included in each of the figures have also been separately saved as a comma-separated values file, `yearly_average_anomaly.csv`, and may be retrieved via <https://github.com/USEPA/ri_wq_trends/blob/master/data/yearly_average_anomaly.csv>.

## Water quality trend analysis

@@

Median annual temperature anomalies in lakes and reservoirs appear to be increasing as the slope is greater than the threshold of 0.02 we identified, the p-value suggests some statistical support for a positive trend (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`), and the majority of years with median temperature greater than the long-term median are occurring in the second half of the time period (`r fig_num(name = "temp_ri", caption = "Twenty year trend for median lake temperature anomaly in Rhode Island. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians. Black line is the linear fit for year and median anomaly.", display = "cite")`, `r tab_num(name = "trend_summ", caption = 'Summary of long term water quality trends in the URIWW and LAGOSNE data for termperature (Temp.) chlorophyll (Chl.), total nitrogen (TN), total phosphorus (TP), and the molar nitrogen to phosphorus ratio (N:P). The "++" indicate a positive trend, the "+" indicates a weak positive trend, and the "0" indicates no trend.', display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`).  The slope of `r signif(ww_chla_rl$slope,2)` μg/L is greater than our threshold of 0.035 μg/L and the above-average years have mostly occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Twenty year trend for median chlorophyll anomaly.  Panel A. URI Watershed Watch yearly chlorophyll anomalies.  Panel B. LAGOSNE yearly chlorophyll anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians. Black line is the linear fit for year and median anomaly.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).

@@

In general, there was little evidence to suggest broad regional changes in chlorophyll *a* as it showed a very weak positive trend, slight statistical support and above average years spread evenly throughout the time period (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Total nitrogen showed a slight decreasing trend with weak statistical support and slope less than the 3.5 μg/L that would result in a trophic state change over a century, thus there is little support for a meaningful trend in TN at the regional scale (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Furthermore, The last two years for which LAGOSNE had data on total nitrogen were from a relatively smaller number of sites (`r fig_num("year_n", display = "cite")`) and those values may not be representative.  Total phosphorus showed a very small increasing trend and there is statistical support for that trend.  The slope however suggests little support for an ecological meaningful change in total phosphorus (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B., `r tab_num(name = "", display = "cite")`). Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "
", display = "cite")`) as none of our criteria for a trend were met.  Although, there appears to be a non-linear trend that suggests increasing TN:TP early in the time period and decreasing TN:TP later, we feel we cannot effectively evaluate this though as total nitrogen measurements were rare in 2011 (`r fig_num("year_n", display = "cite")`) and sites that had both nitrogen and phosphorus were even rarer.  This low median anomaly should be interpreted with this caveat in mind.   Taken together we feel our results largely match the findings of Oliver et al. [-@Oliver_2017] that there is stasis in lake nutrients and chlorophyll within the LAGOSNE region.

@@

```{r context, echo=FALSE, results="asis"}
#knitr::kable(context_summ, format = "pipe")
flextable(context_summ) %>%
  width(width = 1) %>%
  fontsize(size = 11, part = "all") %>%
  bold(part = "header") %>%
  font(fontname = "Times New Roman", part = "all") %>%
  autofit()
```
`r tab_num("context_summ")`

@@

```{r ww_params, echo=FALSE, results="asis"}
#knitr::kable(ww_param_ranges, format = "pipe")
flextable(ww_param_ranges) %>%
  width(width = 1) %>%
  fontsize(size = 11, part = "all") %>%
  bold(part = "header") %>%
  font(fontname = "Times New Roman", part = "all") %>%
  autofit()
```
`r tab_num("ww_params")`

@@

```{r lagos_params, echo=FALSE, results="asis"}
#knitr::kable(lagos_param_ranges, format = "pipe")
flextable(lagos_param_ranges) %>%
  width(width = 0.5) %>%
  fontsize(size = 11, part = "all") %>%
  bold(part = "header") %>%
  font(fontname = "Times New Roman", part = "all") %>%
  autofit()
```
`r tab_num("lagos_params")`

@@

```{r ts_table, echo=FALSE, results="asis"}
#knitr::kable(ts_table, format = "pipe")
flextable(ts_table) %>%
  width(width = 1.2) %>%
  fontsize(size = 11, part = "all") %>%
  bold(part = "header") %>%
  font(fontname = "Times New Roman", part = "all") %>%
  autofit()
```
`r tab_num("ts_table")`

@@

trend_summ_df <- tibble(Source = c("URIWW", "LAGOSNE"), 
                        Temp. = c("++", ""), Chl. = c("++", "0"), 
                        `TN` = c("0", "0"), 
                        `TP`= c("0", "+"), 
                        `N:P` = c("0", "0"))

#knitr::kable(trend_summ_df, format = "pipe")
flextable(trend_summ_df) %>%
  width(width = 1) %>%
  fontsize(size = 11, part = "all") %>%
  bold(part = "header") %>%
  font(fontname = "Times New Roman", part = "all") %>%
  autofit()
```
`r tab_num("trend_summ")`
